First steps in synthetic algebraic geometry

### 2017-03-15 snippets/choice-site.tex
On the proper choice of a suitable site for constructing the big Zariski topos

\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\FF}{\mathbb{F}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\A}{\mathcal{A}}
\renewcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\G}{\mathcal{G}}
\renewcommand{\H}{\mathcal{H}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\L}{\mathcal{L}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\T}{\mathcal{T}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\GG}{\mathbb{G}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\qqq}{\mathfrak{q}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\nnn}{\mathfrak{n}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\HOM}{\mathcal{H}\mathrm{om}}
\newcommand{\id}{\mathrm{id}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\placeholder}{\underline{\quad}}
\newcommand{\ul}[1]{\underline{#1}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Grp}{\mathrm{Grp}}
\newcommand{\Vect}{\mathrm{Vect}}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\PSh}{\mathrm{PSh}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\Et}{\mathrm{\acute{E}t}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Mod}{\mathrm{Mod}}
\newcommand{\Alg}{\mathrm{Alg}}
\newcommand{\Ring}{\mathrm{Ring}}
\newcommand{\LRL}{\mathrm{LRL}}
\newcommand{\pt}{\mathrm{pt}}
\newcommand{\tors}{\mathrm{tors}}
\DeclareMathOperator{\Spec}{Spec}
\newcommand{\QcohSpec}[2]{\mathrm{Spec}^{\mathrm{qcoh}}_{#1}{#2}}
\newcommand{\RelSpec}[2]{\mathrm{RelSpec}_{#1}{#2}}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\colim}{colim}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\Ann}{Ann}
\DeclareMathOperator{\Int}{int}
\DeclareMathOperator{\Clos}{cl}
\DeclareMathOperator{\Kernel}{ker}
\DeclareMathOperator{\supp}{supp}
\newcommand{\Ass}{\mathrm{Ass}}
\newcommand{\Open}{\T}
\newcommand{\?}{\,{:}\,}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\Ll}{\vcentcolon\Longleftrightarrow}
\newcommand{\notat}[1]{{!#1}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\lhra}{\ensuremath{\lhook\joinrel\relbar\joinrel\rightarrow}}
\newcommand{\hra}{\hookrightarrow}
\newcommand{\brak}[1]{{\llbracket{#1}\rrbracket}}
\newcommand{\sdense}{{\widehat\Box}}
\newcommand{\sdenseother}{\Box}
\newcommand{\ie}{i.\,e.\@\xspace}
\newcommand{\eg}{e.\,g.\@\xspace}
\newcommand{\vs}{vs.\@\xspace}
\newcommand{\resp}{resp.\@\xspace}
\newcommand{\inv}{inv.\@}
\newcommand{\notnot}{\emph{not not}\xspace}
\newcommand{\affl}{\ensuremath{{\ul{\AA}^1_S}}\xspace}
\newcommand{\afflx}{\ensuremath{{\ul{\AA}^1_X}}\xspace}
\newcommand{\affla}{\ensuremath{{\ul{\AA}^1_{\Spec A}}}\xspace}
\newcommand{\xra}{\xrightarrow}

@@

\begin{prop}The following properties of~$\affl$, as seen from the internal
point of view of~$\Zar(S)$, are consequences of~$\affl$ satisfying the
quasicoherence condition (and~$\affl$ not being the trivial ring):
% Adjust paragraph below if numbering changes.
\begin{enumerate}
\item $\affl$ is a field in the sense that
$\forall x\?\affl\_ \neg(x = 0) \Rightarrow \speak{$x$ \inv}$.
\item $\affl$ is not a reduced ring: $\neg \forall x\?\affl\_ (\bigvee_{n \geq
0} x^n = 0) \Rightarrow x = 0.$
\item Any function~$\affl \to \affl$ is given by a unique polynomial
in~$\affl[T]$.
\item Any monic polynomial~$p \in \affl[T]$ of degree at least one does \notnot have a
zero.
\end{enumerate}\end{prop}

\begin{proof}\begin{enumerate}
\item Let $x\?\affl$ such that~$\neg(x=0)$. We consider the quasicoherence
condition for the finitely presented~$\affl$-algebra~$A \defeq \affl/(x)$.
Since~$\Spec(A) \cong \brak{x=0} = \brak{\bot} = \emptyset$, the condition says
that the canonical homomorphism
\[ \affl/(x) \longrightarrow [\emptyset, \affl] \]
is an isomorphism. Since its codomain is the zero algebra, so is~$\affl/(x)$.
Therefore~$1 \in (x)$, that is,~$x$ is invertible.
\item Assume that~$\affl$ is reduced. Then the set~$\Delta \defeq \{
\varepsilon \in \affl \,|\, \varepsilon^2 = 0 \}$ is equal to~$\{ 0 \}$.
By the quasicoherence criterion applied to the finitely
presented~$\affl$-algebra~$A \defeq \affl[T]/(T^2)$, the canonical map
\[ \affl[T]/(T^2) \longrightarrow [\Spec(\affl[T]/(T^2)), \affl] \cong
  [\Delta, \affl] \cong \affl \]
is an isomorphism. It maps~$[T]$ to zero (the value of~$T$ at~$0 \in \Delta$).
Thus~$T \in (T^2)$ and therefore~$1 = 0$ in~$\affl$. This is a contradiction.
\item Immediate by considering the quasicoherence condition for the finitely
presented~$\affl$-algebra~$A \defeq \affl[T]$ and noticing that~$\Spec(A) \cong
\affl$.
\item Let~$p \in \affl[T]$ be a monic polynomial of degree at least one. Assume
that~$p$ doesn't have a zero in~$\affl$. Then the spectrum of~$A \defeq
\affl[T]/(p)$ is empty. The quasicoherence condition for~$A$ therefore implies
that~$\affl[T]/(p)$ is zero. This means that~$p$ is invertible in~$\affl[T]$.
A basic lemma in commutative algebra (whose standard proof is constructive)
then implies that with the exception of the constant term in~$p$, all
coefficients are nilpotent. This contradicts the assumption that~$p$ is monic
of degree at least one.\qedhere
\end{enumerate}\end{proof}

In classical logic, the first two properties would directly contradict each
other; only an intuitionistic context allows for fields which are not reduced.
Also the third statement cannot be satisfied in classical logic: for infinite
fields the existence part fails and for finite fields the uniqueness part
fails.

That~$\affl$ is not reduced, irrespective of the reducedness of the base
scheme~$S$, should not come as a surprise: Reducedness is not stable under base
change, but all statements of the internal language of~$\Zar(S)$ are.
If~$\affl$ was reduced, then all~$S$-schemes (at least those contained in the
site used to construct~$\Zar(S)$) would be reduced as well. In contrast, the
structure sheaf~$\O_S$ is reduced from the point of view of the little Zariski
topos if and only if~$S$ is reduced (Proposition~\ref{prop:reduced-ring}).

>>>>

\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\FF}{\mathbb{F}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\A}{\mathcal{A}}
\renewcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\G}{\mathcal{G}}
\renewcommand{\H}{\mathcal{H}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\L}{\mathcal{L}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\T}{\mathcal{T}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\renewcommand{\S}{\mathcal{S}}
\renewcommand{\U}{\mathcal{U}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\GG}{\mathbb{G}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\qqq}{\mathfrak{q}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\nnn}{\mathfrak{n}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\HOM}{\mathcal{H}\mathrm{om}}
\newcommand{\id}{\mathrm{id}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\placeholder}{\underline{\quad}}
\newcommand{\ul}[1]{\underline{#1}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Grp}{\mathrm{Grp}}
\newcommand{\Vect}{\mathrm{Vect}}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\PSh}{\mathrm{PSh}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\Et}{\mathrm{\acute{E}t}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\Mod}{\mathrm{Mod}}
\newcommand{\Alg}{\mathrm{Alg}}
\newcommand{\Ring}{\mathrm{Ring}}
\newcommand{\LRL}{\mathrm{LRL}}
\newcommand{\pt}{\mathrm{pt}}
\newcommand{\tors}{\mathrm{tors}}
\newcommand{\lfp}{\mathrm{lfp}}
\newcommand{\fp}{\mathrm{fp}}
\DeclareMathOperator{\Spec}{Spec}
\newcommand{\QcohSpec}[2]{\mathrm{Spec}^{\mathrm{qcoh}}_{#1}{#2}}
\newcommand{\RelSpec}[2]{\mathrm{RelSpec}_{#1}{#2}}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\colim}{colim}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\Ann}{Ann}
\DeclareMathOperator{\Int}{int}
\DeclareMathOperator{\Clos}{cl}
\DeclareMathOperator{\Kernel}{ker}
\DeclareMathOperator{\supp}{supp}
\newcommand{\Ass}{\mathrm{Ass}}
\newcommand{\Open}{\T}
\newcommand{\?}{\,{:}\,}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\Ll}{\vcentcolon\Longleftrightarrow}
\newcommand{\notat}[1]{{!#1}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\lhra}{\ensuremath{\lhook\joinrel\relbar\joinrel\rightarrow}}
\newcommand{\hra}{\hookrightarrow}
\newcommand{\brak}[1]{{\llbracket{#1}\rrbracket}}
\newcommand{\sdense}{{\widehat\Box}}
\newcommand{\sdenseother}{\Box}
\newcommand{\ie}{i.\,e.\@\xspace}
\newcommand{\eg}{e.\,g.\@\xspace}
\newcommand{\vs}{vs.\@\xspace}
\newcommand{\resp}{resp.\@\xspace}
\newcommand{\inv}{inv.\@}
\newcommand{\notnot}{\emph{not not}\xspace}
\newcommand{\affl}{\ensuremath{{\ul{\AA}^1_S}}\xspace}
\newcommand{\afflx}{\ensuremath{{\ul{\AA}^1_X}}\xspace}
\newcommand{\affla}{\ensuremath{{\ul{\AA}^1_{\Spec A}}}\xspace}
\newcommand{\xra}{\xrightarrow}

@@

Unlike with the construction of the little Zariski topos, set-theoretical
issues of size arise when constructing the big Zariski topos. These can be
solved in several different manners, yielding toposes which are not equivalent,
and actually differ in some important aspects, but otherwise enjoy very similar
properties.

Some authors construct the big Zariski topos of~$S$ as the topos of
sheaves over the site~$\Sch/S$ of all schemes over~$S$. This option is quite
attractive since the Yoneda functor~$\Sch/S \to \Sh(\Sch/S)$, which sends
an~$S$-scheme to its functor of points, is fully faithful. However,
since~$\Sch/S$ is not essentially small, forming the sheaf topos is not
possible in plain Zermelo--Fraenkel set theory.
Since it's still possible to meaningfully speak of individual
functors~$(\Sch/S)^\op \to S$, we can attach a Kripke--Joyal semantics
to~$\Sh(\Sch/S)$, as long as one keeps in mind that~$\Sh(\Sch/S)$ might
not contain a subobject classifier.

We could also assume the existence of a Grothendieck universe~$\U$
containing~$S$ and construct~$\Zar(S)$ as the topos of sheaves over the
small site~$\Sch_\U/S$, the category of~$S$-schemes contained in~$U$.
By the \emph{comparison lemma}~\ref{XXX}, we could also construct~$\Zar(S)$ as
the topos of sheaves over~$\Aff_\U/S$, the category of~$S$-schemes in~$U$ which
are affine (as absolute schemes), and obtain an equivalent topos.
In this case, the Yoneda functor~$\Sch/S \to \Zar(S)$ might not be faithful,
but the restricted Yoneda functor~$\Sch_\U/S \to \Zar(S)$ is.

The Stacks Project proposes a more nuanced approach, namely expanding a given
set~$M_0$ of schemes containing~$S$ to a superset~$M$ which is closed (up to
isomorphism) under several constructions~\cite[Tag~000H]{stacks-project}: fiber
products, countable coproducts, domains of open and closed immersions and of
morphisms of finite type, spectra of local rings~$\O_{X,x}$, spectra of residue
fields, and others. The Stacks Project then defines~$\Zar(S)$
as~$\Sh(\Sch_M/S)$, where~$\Sch_M/S$ is the small category of~$S$-schemes
in~$M$, or equivalently as~$\Sh(\Aff_M/S)$. This approach has the advantage
that one doesn't have to assume the existence of a Grothendieck universe.

From a topos-theoretical point of view, it's natural to settle for an even
more parsimonious site: the site~$(\Sch/S)_\lfp$ consisting of the~$S$-schemes
which are locally of finite presentation over~$S$, or equivalently the
essentially small site~$(\Aff/S)_\lfp$ of the~$S$-schemes which are locally of
finite presentation over~$S$ and affine (as absolute schemes).\footnote{It's
not reasonable to restrict to the even smaller site consisting of the finitely
presented~$S$-schemes, since open immersions can fail to be finitely presented.
We want the site used to construct~$\Zar(S)$ to be closed under open immersions
for various reasons, for instance to facilitate a comparison with the little
Zariski topos~$\Sh(S)$, whose site does contain all open subsets of~$S$.
Furthermore, since a finitely presented~$S$-scheme might not admit an open
covering by finitely presented~$S$-schemes which are affine (as absolute
schemes), the toposes~$\Sh((\Sch/S)_\fp)$ and~$\Sh((\Aff/S)_\fp)$ can differ.}

In the special case that~$S = \Spec(A)$ is affine, this site is the
dual of the category of finitely presented~$A$-algebras; in this case the
topos-theoretic points of the resulting topos are precisely the local~$A$-algebras,
and moreover, the resulting topos is the classifying topos of the theory of
local~$A$-algebras, such that for any Grothendieck topos~$\E$, geometric
morphisms~$\E \to \Sh((\Aff/S)_\lfp)$ correspond to local~$A$-algebras internal
to~$\E$.

In contrast, the toposes arising when using the larger sites have categories of
points which contain further objects in addition to all local~$A$-algebras; and
no simple description of the theory they classify is known.

A disadvantage of employing the sites with the presentation condition is that
the Yoneda functor is only fully faithful when restricted to~$(\Sch/S)_\lfp$.
For instance, in the case~$S = \Spec(\ZZ)$, the schemes~$\Spec(\QQ)$ and the
empty scheme have isomorphic functors of points, whereby~$\Spec(\QQ)$ looks
like the empty set from the internal point of view.\footnote{XXX explicit proof}

In the following, we try to encompass XXX point out XXX
